---
title: "How Gaming Affects Well-being"
author: "Sandy, Fredy, Benjamin"
date: "2026-04-19"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#install.packages(c("tidyverse", "readxl", "ggplot2", "ggridges"))
library(readr)
library(dplyr)
library(tidyverse)
library(readxl)
library(ggridges)

# Code Setup
gamingA_raw <- read.csv("GamingStudy_data.csv", stringsAsFactors = FALSE)
names(gamingA_raw)
str(gamingA_raw)

# Keep variables we want
gamingA <- gamingA_raw %>%
  select(
    Hours, streams, Game, Platform, whyplay, Playstyle,
    GAD_T, SWL_T, SPIN_T,
    Narcissism, Residence, Age, Work, Degree, Gender
  ) %>%
  drop_na(Hours, GAD_T, SWL_T, SPIN_T)

# Make sure numeric columns are numeric
gamingA <- gamingA %>%
  mutate(
    Hours = as.numeric(Hours),
    streams = as.numeric(streams),
    GAD_T = as.numeric(GAD_T),
    SWL_T = as.numeric(SWL_T),
    SPIN_T = as.numeric(SPIN_T),
    Narcissism = as.numeric(Narcissism),
    Age = as.numeric(Age)
  )

# Remove impossible or extreme values if needed
gamingA <- gamingA %>%
  filter(
    Age >= 18,
    Age < 66,
    Hours >= 0,
    Hours <= 112
  )

# Create a total gaming variable
gamingA <- gamingA %>%
  mutate(total_gaming = Hours + ifelse(is.na(streams), 0, streams))

# Create gaming groups
gamingA <- gamingA %>%
  mutate(
    gaming_level = case_when(
      total_gaming < 5 ~ "Low",
      total_gaming < 15 ~ "Moderate",
      total_gaming < 30 ~ "High",
      TRUE ~ "Extreme"
    ),
    gaming_level = factor(gaming_level,
                          levels = c("Low", "Moderate", "High", "Extreme"))
  )
```

## Introduction

**Domain Question:**
How does gaming affect our well-being?

*Sub-questions:*
Does high gaming intensity correlate with anxiety (GAD)?

How does gaming affect social interaction & social anxiety (SPIN)?

How does gaming affect Life Satisfaction (SWL)?

## Data Overview
* Authentic data
* Surveyed individuals
* Thousands of entries (Large Dataset)

A look at our sample's gaming habits.

```{r}
# Gaming Hours
ggplot(gamingA, aes(x = Hours)) +
  geom_histogram(binwidth = 5, fill = "gray70", color = "black") +
  labs(
    title = "Distribution of Weekly Gaming Hours",
    x = "Hours Played per Week",
    y = "Number of Gamers"
  ) +
  theme_classic()
```

## Psychological Effects of Gaming

**Anxiety**

Does more gaming mean more anxiety?

```{r}
# Mental health groups
ggplot(gamingA, aes(x = gaming_level, y = GAD_T, fill = gaming_level)) +
  geom_boxplot() +
  labs(
    title = "Anxiety Scores by Gaming Level",
    x = "Gaming Intensity Level",
    y = "Anxiety Score",
    fill = "Gaming Level"
  ) +
  theme_classic()
```

**Average Anxiety**

```{r}
# Creating a 10 hour group
gamingA <- gamingA %>%
  mutate(hours_group = ntile(Hours, 10))

# Mena summaries 
agg <- gamingA %>%
  group_by(hours_group) %>%
  summarise(
    anxiety = mean(GAD_T, na.rm = TRUE),
    life = mean(SWL_T, na.rm = TRUE),
    social = mean(SPIN_T, na.rm = TRUE)
  )

# plotting trends
ggplot(agg, aes(x = hours_group, y = anxiety)) +
  geom_col(fill = "gray60") +
  labs(
    title = "Average Anxiety by Gaming Hour Group",
    x = "Gaming Hour Group",
    y = "Average Anxiety Score"
  ) +
  theme_classic()

```

**Anxiety by Residence**

Anxiety may not be caused by gaming alone.

```{r}
# Residence
top_res <- gamingA %>%
  count(Residence, sort = TRUE) %>%
  slice_head(n = 8) %>%
  pull(Residence)

gaming_res <- gamingA %>%
  filter(Residence %in% top_res)

# Using residence to compare
ggplot(gaming_res, aes(x = Residence, y = GAD_T, fill = Residence)) +
  geom_boxplot() +
  labs(
    title = "Anxiety by Residence",
    x = "Residence",
    y = "Anxiety Score"
  ) +
  theme_classic() +
  theme(legend.position = "none")

gamingA <- gamingA %>%
  mutate(high_anxiety = ifelse(GAD_T >= 8, 1, 0))

gamingA %>%
  group_by(Residence) %>%
  summarise(avg_anxiety = mean(GAD_T)) %>%
  arrange(desc(avg_anxiety)) %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(Residence, avg_anxiety), y = avg_anxiety)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top Countries by Anxiety",
    x = "Residence",
    y = "Anxiety Score")
```


## Why do People play Games?

**What are the primary drivers for gaming?**

```{r}
gamingA <- gamingA %>%
  mutate(
    whyplay_clean = case_when(
      
      str_detect(whyplay, "fun") ~ "Fun",
      
      str_detect(whyplay, "relax") ~ "Relaxation",
      
      str_detect(whyplay, "improv") ~ "Improvement",
      
      str_detect(whyplay, "win") ~ "Competition",
      
      TRUE ~ "Other"
    )
  )

ggplot(gamingA, aes(x = whyplay_clean, fill = whyplay_clean)) +
  geom_bar() +
  labs(
    title = "Why People Play Games",
    x = "Gaming Motivation",
    y = "Number of Gamers",
    fill = "Gaming Motivation"
  ) +
  theme_classic()
```


Why you play and who you play with might matter more than how long you play.


```{r}
gamingA <- gamingA %>%
  mutate(
    Playstyle_clean = case_when(
      
      str_detect(Playstyle, "Single") ~ "Singleplayer",
      
      str_detect(Playstyle, "offline") ~ "Offline Multiplayer",
      
      str_detect(Playstyle, "real life friends") ~ "Online with Friends",
      
      str_detect(Playstyle, "strangers") ~ "Online with Strangers",
      
      str_detect(Playstyle, "acquaintances") ~ "Online Mixed",
      
      TRUE ~ "Other"
    )
  )

ggplot(gamingA, aes(x = Playstyle_clean, y = GAD_T, fill = Playstyle_clean)) +
  geom_boxplot() +
  labs(
    title = "Anxiety Score by Gaming Playstyle",
    x = "Playstyle",
    y = "Anxiety Score",
    fill = "Playstyle"
  ) +
  theme_classic()

ggplot(gamingA, aes(x = whyplay_clean, y = GAD_T, fill = whyplay_clean)) +
  geom_boxplot() +
  labs(
    title = "Anxiety Score by Gaming Motivation",
    x = "Gaming Motivation",
    y = "Anxiety Score",
    fill = "Gaming Motivation"
  ) +
  theme_classic()
```


## Social Effects of Gaming

We compared social anxiety scores across different playstyles to see how who you play with affects your gaming experience.

```{r}
# Social anxiety
ggplot(gamingA, aes(x = gaming_level, y = SPIN_T, fill = gaming_level)) +
  geom_boxplot() +
  labs(
    title = "Social Anxiety by Gaming Level",
    x = "Gaming Level",
    y = "Social Anxiety Score",
    fill = "Gaming Motivation"
  ) +
  theme_classic()

ggplot(agg, aes(x = hours_group, y = social)) +
  geom_col(fill = "gray60") +
  labs(
    title = "Average Social Anxiety by Gaming Hour Group",
    x = "Gaming Hour Group",
    y = "Average Social Anxiety Score"
  ) +
  theme_classic()

gamingA %>%
  group_by(Playstyle_clean) %>%
  summarise(avg_spin = mean(SPIN_T, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(Playstyle_clean, avg_spin), y = avg_spin, fill = Playstyle_clean)) +
  geom_col() +
  scale_fill_brewer(palette = "Pastel1") +
  coord_flip() +
  labs(
    title = "Social Anxiety by Playstyle",
    x = "Playstyle",
    y = "Social Anxiety Score",
    fill = "Playstyle"
  ) +
  theme_classic() +
  theme(legend.position = "none")
```


## Life Satisfaction

**How does gaming affects happiness?**

```{r}
# Why people play
ggplot(gamingA, aes(x = SWL_T, y = reorder(whyplay_clean, SWL_T, FUN = median), fill = whyplay_clean)) +
  geom_density_ridges(alpha = 0.8, color = "white") +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Life Satisfaction by Motivation for Playing",
    x = "Life Satisfaction Score",
    y = "Gaming Motivation",
    fill = "Gaming Motivation"
  ) +
  theme_ridges() 

# Playstyle
gamingA %>%
  group_by(Playstyle_clean) %>%
  summarise(avg_spin = mean(SPIN_T, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(Playstyle_clean, avg_spin), y = avg_spin, fill = Playstyle_clean)) +
  geom_col() +
  scale_fill_brewer(palette = "Pastel1") +
  coord_flip() +
  labs(
    title = "Social Anxiety by Playstyle",
    x = "Playstyle",
    y = "Social Anxiety Score",
    fill = "Playstyle"
  ) +
  theme_classic()

ggplot(gamingA, aes(x = whyplay_clean, y = SWL_T, fill = whyplay_clean)) +
  geom_boxplot() +
  labs(title = "Life Satisfaction by Motivation for Playing",
    x = "Gaming Motivation",
    y = "Life Satisfaction Score",
    fill = "Gaming Motivation"
    )
```

## Demographic Trends

**Who is playing and how does it affect them?**

Is high-intensity gaming more stressful for older or younger people?

```{r}
gamingA <- gamingA %>%
  mutate(age_bracket = cut(Age, breaks = c(18, 25, 35, 45, 66), 
                           labels = c("18-25", "26-35", "36-45", "46+")))
gamingA %>%
  filter(!is.na(age_bracket), !is.na(gaming_level)) %>%
  group_by(age_bracket, gaming_level) %>%
  summarise(avg_anxiety = mean(GAD_T, na.rm = TRUE)) %>%
  ggplot(aes(x = gaming_level, y = age_bracket, fill = avg_anxiety)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "turbo") +
  labs(
    title = "Age Group vs Gaming Level",
    x = "Gaming Level",
    y = "Age Bracket",
    fill = "Avg Anxiety Score"
  ) + theme_minimal()

```

## Key Findings


**Anxiety:** Shows an upward trend in "Extreme" gaming levels.

**Social Context:** Playing with strangers correlates with higher Social Anxiety compared to playing with friends.

**Motivation:** Players gaming for "Relaxation" and "Fun" report higher Life Satisfaction than those gaming for "Competition."

**Final Thoughts:** Gaming's effect on well-being depends heavily on how and why you play.

## What does it mean?

* Well-being remains stable through moderate gaming

But in the Extreme category (>30 hours/week) anxiety begins to rise.


* Social Anxiety isn't tied to how much you play but who you play with 

Playing with friends results in lower social anxiety than playing with strangers or alone.


* Gamers motivated by fun and relaxation are more satisfied

Those driven primarily by competition or escapism experience more anxiety.


* Gaming doesn't affect everyone the same way

Our data suggests the well-being of a gamer is determined by the balance between their gaming hours and their social setting. 

## What We Suggest

If you're playing games, you should pay attention to:

* How much you play

* Who you're playing with

* Why you're playing

If all three of these reasons align with actions that don't bring you stress then your overall well-being will increase rather than decrease.
