---
title: "The Relationship Between Gaming & Mental Health"
author: "Sandy, Fredy, Benjamin"
date: "2026-04-14"
output:
  slidy_presentation: default
  ioslides_presentation: default
---
```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(tidyverse)
library(readxl)
library(patchwork)

#uncomment to read csv
#df <- read.csv("gaming_mental_health_10M_40features.csv")

load("smallerdata.Rdata")

#created a sub dataset based on the variables we need. Use gaming_subdata from now on.
gaming_subdata <- gaming_data %>%
  select(age, gender, income, bmi, daily_gaming_hours, weekly_sessions, years_gaming, eye_strain_score, back_pain_score, sleep_hours, exercise_hours, happiness_score, aggression_score, loneliness_score, stress_level, anxiety_score, depression_score, social_interaction_score, addiction_level, toxic_exposure, violent_games_ratio, parental_supervision, microtransactions_spending, competitive_rank)

#renamed gaming_data to gaming for seamless integration into my code
gaming <- gaming_subdata %>% 
  sample_n(5000)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
physical_data <- gaming %>%
  select(daily_gaming_hours, bmi, exercise_hours, eye_strain_score, back_pain_score) %>%
  drop_na()

```

## Introduction

**Domain Question:**
How does gaming affect our well-being?

*Sub-question:*
What are the physical, psychological, & social effects of gaming?

## Data Overview
* Synthetic dataset
* Randomized data
* Thousands of entries (Large dataset)

## Physical Effects of Gaming

Back Pain

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
# adding questions: Question: Does gaming increase back pain?
ggplot(physical_data, aes(x = daily_gaming_hours, y = back_pain_score)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs Back Pain") +
  theme_minimal()
# answer to the question: No. Back pain slightly decreases as gaming hours increase, but the effect seems to be very small and likely not meaningful
#Other possible answer: Surprisingly, the data shows a slight decrease in back pain with more gaming. However, the change is very small, so this relationship is likely weak or due to noise in the data rather than a real effect.
```

## Physical Effects of Gaming

Eye Strain

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
# Does gaming affect eye strain?
ggplot(physical_data, aes(x = daily_gaming_hours, y = eye_strain_score)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs Eye Strain") +
  theme_minimal()
# Yes. Eye strain increases as gaming hours increase, most likely due to prolonged screen time. 
```

## Physical Effects of Gaming

Exercise Hours

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
# Does more gaming reduce exercise?
ggplot(physical_data, aes(x = daily_gaming_hours, y = exercise_hours, fill = daily_gaming_hours)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs Exercise Hours") +
  theme_minimal()
#Yes. Exercise hours slightly decrease as gaming hours increase, indicating a more sedentary lifestyle.
```

## Physical Effects of Gaming

BMI

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Does gaming affect BMI?
ggplot(physical_data, aes(x = daily_gaming_hours, y = bmi)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs BMI") +
  theme_minimal()
#BMI slightly increases with more gaming, but the effect is very small, suggesting a weak relationship.
```

## Psychological Effects of Gaming

Addiction

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Does more gaming increase addiction levels? Is there a postive realtion of addiction and gaming hours?
ggplot(gaming, aes(x = daily_gaming_hours, y = addiction_level)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Gaming Hours vs Addiction Level",
         x = "Daily Gaming Hours",
         y = "Addiction Level") +
  theme_minimal()
# Yes. Addiction levels increase strongly with gaming hours, showing a clear positive relationship.
```

## Psychological Effects of Gaming

Stress

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
# Stress slightly decreases as gaming hours increase, suggesting gaming may help reduce stress, though the effect is small.
ggplot(gaming, aes(x = daily_gaming_hours, y = stress_level)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs Stress Level") +
  theme_minimal()
```


## Psychological Effects of Gaming

Anxiety

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Anxiety increases slightly with more gaming, but the relationship is weak.
ggplot(gaming, aes(x = daily_gaming_hours, y = anxiety_score)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs Anxiety Score") +
  theme_minimal()
```

## Psychological Effects of Gaming

Depression

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Depression slightly decreases as gaming hours increase, though the effect is minimal.
ggplot(gaming, aes(x = daily_gaming_hours, y = depression_score)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Gaming Hours vs Depression Score") +
  theme_minimal()
```

## Psychological Effects of Gaming

Happiness

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Happiness increases with gaming hours, suggesting gaming may have positive emotional effects.
ggplot(gaming, aes(x = daily_gaming_hours, y = happiness_score)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Gaming Hours vs Happiness Score",
         x = "Daily Gaming Hours",
         y = "Happiness Score") +
  theme_minimal()
```

## Gaming's Effect on Mental Health

Overall psychological toll of gaming on mental health 

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
gaming <- gaming %>%
  mutate(
    mental_health_score = (stress_level + anxiety_score + depression_score) / 3
  )

gaming <- gaming %>%
  mutate(
    gaming_group = case_when(
      daily_gaming_hours < 2 ~ "Low",
      daily_gaming_hours < 5 ~ "Moderate",
      TRUE ~ "High"
    )
  )

gaming %>%
  group_by(gaming_group) %>%
  summarise(
    avg_stress = mean(stress_level, na.rm = TRUE),
    avg_anxiety = mean(anxiety_score, na.rm = TRUE),
    avg_depression = mean(depression_score, na.rm = TRUE)
  )

ggplot(gaming, aes(x = gaming_group, y = mental_health_score, fill = gaming_group)) +
  geom_boxplot() +
  labs(
    title = "Mental Health by Gaming Level",
    x = "Gaming Level",
    y = "Mental Health Score"
  ) +
  theme_minimal()

```

## Social Effects of Gaming

Loneliness

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Loneliness increases as gaming hours increase, indicating potential social isolation.
ggplot(gaming, aes(x = daily_gaming_hours, y = loneliness_score)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Gaming Hours vs Loneliness Score",
         x = "Daily Gaming Hours",
         y = "Loneliness Score") +
  theme_minimal()
```

## Social Effects of Gaming

Social Interaction

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
#Social interaction slightly decreases as gaming hours increase, suggesting reduced real-world interaction.
ggplot(gaming, aes(x = daily_gaming_hours, y = social_interaction_score)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Gaming Hours vs Social Interaction Score",
         x = "Daily Gaming Hours",
         y = "Social Interaction Score") +
  theme_minimal()
```

## Other Metrics

Data By Age

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
data_by_age <- gaming %>%  
  mutate(age_group = case_when(
    age < 18 ~ "Under 18",
    age >= 18 & age <= 25 ~ "18-25",
    age > 25 ~ "25+"
  )) %>% 
  group_by(age_group) %>% 
  summarize(
    avg_addiction = mean(addiction_level, na.rm = TRUE),
    .groups = 'drop'
  )

ggplot(data_by_age, aes(x = age_group, y = avg_addiction)) +
  geom_col() + 
  labs(title = "Average Addiction by Age",
                    x = "Age",
                    y = "Average Addiction Score") +
  theme_minimal()
```

## Other Metrics

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
data_by_gender <- gaming %>% 
  group_by(gender) %>% 
  summarize(
    avg_addiction = mean(addiction_level, na.rm = TRUE),
    .groups = 'drop'
  )

ggplot(data_by_gender, aes(x = gender, y = avg_addiction)) +
  geom_col() + 
  labs(title = "Average Addiction by Gender",
                    x = "Gender",
                    y = "Average Addiction Score") +
  theme_minimal()
```

## Drawbacks

**Suspicious Reports**

* Underreporting - High hours, low addiction score

* Overreporting - Low hours, high addiction score

If we rule these out, we can focus on consistent behaviors

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4}
gaming <- gaming %>%
  mutate(
    suspicious = case_when(
      daily_gaming_hours > quantile(daily_gaming_hours, 0.75, na.rm = TRUE) &
      addiction_level < quantile(addiction_level, 0.25, na.rm = TRUE) ~ "Underreporting",

      daily_gaming_hours < quantile(daily_gaming_hours, 0.25, na.rm = TRUE) &
      addiction_level > quantile(addiction_level, 0.75, na.rm = TRUE) ~ "Overreporting",

      TRUE ~ "Consistent"
    )
  )
gaming %>%
  count(suspicious)

ggplot(gaming, aes(x = daily_gaming_hours, y = addiction_level, color = suspicious)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Identifying Inconsistent Addiction Reporting",
    x = "Daily Hours Played",
    y = "Addiction Level",
    color = "Classification"
  ) +
  theme_minimal()
```


